---
title: "Water boiling point according to daily pressure"
output: html_document
date: "2023-09-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(ggplot2,dplyr,stringr,lubridate,purrr,cowplot,ncdf4)
```

## Defining boiling point

The boiling point of a substance is the temperature at which the vapor pressure 
of that substance is equal to atmospheric pressure. This relationship can be 
described with the Clausius-Clapeyron Equation.

$ln(\frac{P_1}{P_2}) = \frac{\Delta H_{vap}}{R}(\frac{1}{T_2}-\frac{1}{T_1})$ 

The enthalpy of vaporization of water $\Delta H_{vap} = 40.7 \frac{kJ}{mol}$
and the gas constant $R = 8.3145 \frac{J}{mol K}$

The vapor pressure of water is 1.0 atm at 373 K. We can use this known 
relationship to calculate the daily boiling point of water with the
Clausius-Clapeyron Equation. NCAR records atmospheric pressure at the Foothills
Laboratory in Boulder, Colorado in units of hectopascals 1.0 atm = 1,013.25 hPa. 



## Extracting pressure data

2023 Weather data were extracted for this analysis from the foothills laboratory
(ftp://ftp.eol.ucar.edu/pub/archive/weather/foothills).
Data were processed for atmospheric pressure, recorded every five minutes.

```{r}
file_names <- list.files(path = "weather_data/")
pressure_data <- lapply(paste0("weather_data/", file_names), function(i){
  date <- unlist(str_split(i, "\\."))[2]
  nc_data <- nc_open(i)
  t <- ncvar_get(nc_data, "time")
  pressure <- ncvar_get(nc_data,"pres")
  tibble(date = as_date(date),
         time_sec = t,
         pressure_hPa = pressure)}) %>% 
  reduce(full_join, by = c("date","time_sec","pressure_hPa")) %>% 
  mutate(date = date + seconds(time_sec)) %>% 
  select(-time_sec)

ggplot(data=pressure_data,aes(x=date,y=pressure_hPa)) +
  geom_line() + theme_classic() + labs(y = 'Boiling Point (°F)', x = 'Date')
``` 


## Calculating boiling point according to atmospheric pressure

To calculate the boiling point over time in Boulder, the Clausius-Clapeyron 
equation can be rearranged to 

$T_2 = \frac{1}{\frac{R}{\Delta H_{vap}} (\frac{P_1}{P_2}) + \frac{1}{T_1}}$

By applying this equation to the NCAR dataset, we can determine the variation in
the boiling point of water across 2023 (up to September). 
```{r, fig.width=12}
clausius_func_water <- function(pressure){
  R <- 8.3145/1000
  delta_H <- 40.7
  1/((R/delta_H)*log(1013.25/pressure) + (1/373))
}

pressure_data <- pressure_data %>% 
  mutate(date = date(date)) %>% 
  group_by(date) %>% 
  summarise(pressure_hPa = mean(pressure_hPa)) %>% 
  mutate(boiling_point_K = clausius_func_water(pressure_hPa)) %>% 
  mutate(boiling_point_F = (boiling_point_K - 273.15) * 1.8 + 32)
  

line_plot <- ggplot(data=pressure_data,aes(x=date,y=boiling_point_F)) +
  geom_line() + theme_classic() + ylab('Boiling Point (°F)')
boxplot <- ggplot(data=pressure_data,aes(y=boiling_point_F)) +
  geom_boxplot() + scale_x_discrete() + theme_classic() + 
  ylab('Boiling Point (°F)')
plot_grid(line_plot,boxplot, rel_widths=c(2,1))

mean <- mean(pressure_data$boiling_point_F)
range <- max(pressure_data$boiling_point_F) - min(pressure_data$boiling_point_F)
```

The average calculated boiling point across this time range is `r round(mean, 3)` 
&deg;F, while the difference in minimum and maximum values is `r round(range, 3)`
&deg;F

